
%%% Local Variables:
%%% mode: Xelatex
%%% TeX-master: t
%%% End:

\ctitle{有雾天气下的无人机航拍海上小目标检测方法研究\\
\textcolor{red}{}}

\xuehao{M202374083} \schoolcode{10487}
\csubjectname{计算机技术} \cauthorname{潘颖鹏}
\csupervisorname{孙伟平} \csupervisortitle{副教授}
\defencedate{202X~年~X~月~X~日} \grantdate{}
\chair{}%
\firstreviewer{} \secondreviewer{} \thirdreviewer{}

\etitle{Research on Small Object Detection in Foggy Maritime UAV Aerial Images}
%\textcolor{red}{中英文标题、学科专业、导师姓名正确、一致}}
%\edegree{the Master Degree in Engineering} %（工学硕士）
%\edegree{the Master Degree in Science}     %（理学硕士）
\edegree{the Professional Master Degree}   %(专业学位)
\esubject{Computer Technology}
\eauthor{PAN Yingpeng}
\esupervisor{Assoc. Prof. SUN Weiping }

%定义中英文摘要和关键字
\cabstract{
海上无人机航拍在海事搜救与巡逻中具有重要应用价值，但复杂海洋背景、水面反光以及频发的海雾天气导致航拍图像质量严重退化。雾气引起的对比度衰减和目标边缘模糊，叠加航拍视角下极小的目标尺度，导致现有通用目标检测模型在海上雾天场景中面临严重的误检漏检问题，且成对海上雾天数据集的匮乏进一步限制了去雾检测模型的训练与优化。

针对上述问题，本文以轻量级实时目标检测器 RT-DETR 为基准，提出一种面向海上雾天场景的物理先验引导的轻量级目标检测算法。首先，本文利用单目深度估计与大气散射物理模型，构建了包含多浓度可控参数的合成海上雾天数据集，弥补了训练数据的空白。其次针对微小目标特征提取不足与端侧算力受限问题，构建了基于多维注意力机制的轻量化主干网络。采用较浅层的残差网络以满足实时性要求，并在残差模块中集成无参数三维能量注意力机制与倒残差级联分组自注意力模块，在抑制动态背景噪声的同时显著增强了小目标的底层特征表示能力。针对雾气退化带来的特征模糊问题，提出基于频域解耦与状态空间模型的环境先验提取模块。在特定特征层引入离散小波变换显式解耦高低频特征，将低频特征送入具备线性复杂度全局感受野的状态空间模型中，提取蕴含雾气浓度分布的特征图并进行低频特征恢复。同时设计物理引导特征融合模块，以雾分布特征图作为空间注意力权重将低频恢复特征与高频分支特征进行动态融合，引导在浓雾区域定向补偿高频几何细节，配合分阶段协同训练策略保障了多任务优化的稳定性。

实验结果表明，本文方法在合成有雾AFO数据集上的mAP50达到90.8\%，较基准RT-DETR提升9.3个百分点。该模型将召回率提升至91.5\%，降低了微弱小目标的漏检率，同时将单卡推理帧率维持在82 FPS，兼顾了检测精度与实时性要求。此外，在真实雾天基准RTTS和真实海景SeaDronesSee数据集的跨域测试中，本方法同样展现出较强的的泛化鲁棒性，较好的降低了常规去雾算法引发的边缘过度平滑缺陷。
}

\ckeywords{无人机航拍场景；目标检测；有雾场景；倒残差结构；状态空间模型}



\eabstract{
Unmanned aerial vehicle (UAV) maritime aerial imagery plays an important role in maritime search-and-rescue and patrol missions. However, complex ocean backgrounds, specular reflections on the sea surface, and frequent sea fog severely degrade image quality. The contrast attenuation and edge blurring caused by haze, together with extremely small target scales under aerial viewpoints, lead to pronounced missed detections for generic object detectors in foggy maritime scenarios. Moreover, the lack of paired foggy-clear maritime datasets further limits the training and optimization of dehazing-aware detection models.

To address these issues, this thesis takes the lightweight real-time end-to-end detector RT-DETR as the baseline and proposes a physics-prior-guided lightweight detection framework for foggy maritime UAV scenes. First, a synthetic foggy maritime dataset with controllable multi-level fog density is constructed by combining monocular depth estimation with the atmospheric scattering model, which alleviates the shortage of training data. Second, to cope with insufficient feature representation for tiny targets and the constraints of onboard computation, a multi-dimensional-attention-based lightweight backbone is designed. A shallow residual network is adopted to meet real-time requirements, and a parameter-free 3D energy-based attention module as well as an inverted residual cascaded grouped self-attention module are integrated into residual blocks, effectively suppressing dynamic background noise while enhancing low-level representations of tiny targets. Third, to mitigate haze-induced feature ambiguity, an environmental prior extraction module based on frequency-domain decoupling and state space models is introduced. A discrete wavelet transform is applied at a designated feature level to explicitly separate low- and high-frequency components; the low-frequency features are fed into a state space model with linear complexity and global receptive field to extract haze-density-related feature maps and restore low-frequency structures. In addition, a physics-guided feature fusion module is developed, where the haze-distribution feature map serves as a spatial attention weight to dynamically fuse the restored low-frequency features with high-frequency branch features, enabling targeted compensation of geometric details in dense-haze regions. A stage-wise collaborative training strategy is further employed to improve the stability of multi-task optimization.

Experimental results show that the proposed method achieves an mAP$_{50}$ of 90.8\% on the synthetic foggy AFO dataset, outperforming the RT-DETR baseline by 9.3 percentage points. The recall is improved to 91.5\%, significantly reducing missed detections of weak tiny targets, while maintaining a single-GPU inference speed of 82 FPS, thus balancing accuracy and real-time performance. Furthermore, cross-domain evaluations on the real-world foggy benchmark RTTS and the real maritime dataset SeaDronesSee demonstrate strong generalization and robustness, and the proposed approach alleviates the over-smoothing of edges often introduced by conventional dehazing pipelines.
}

\ekeywords{Drone Aerial Scenes, Object Detection, Foggy Scenes,  Inverted Residual Structure, State Space Model}